Demo arXiv template

### 2022-11-23 paper/paper.qmd
more drafting on the pipeline steps

*constructing time series index should also be encapsulated in the framework*

@fig-pipeline-steps presents the index construction pipeline from the raw data to the final indexes. 
The available intermediate steps include temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, normalising, benchmarking, and simplification. 
These steps do not have to be arranged in sequence and not all the steps are required to present in every index construction. 
Spatial aggregation and temporal processing are operations of the data in space and time. 
Variable transformation, scaling, and dimension reduction concerns operations on the multivariate aspect of the data.
Normalisation, benchmarking, and simplification can be viewed as the post-processing of indexes, which addresses the cross-comparison across different indexes and time, and index communication.

  - divide into pre and post-processing

More steps are involved before the mathematical construction of indexes. 
This includes 1) defining a concept of interest, being useful for decision making and 2) finding the relevant and measurable variables to characterise the concept. 
These choices define the scope of the indexes and the data availability for constructing an index but not the construction itself, hence not included in the pipeline framework. 
The framework proposed in this paper assumes the objective of an index has been defined and relevant variables have been collected and quality controlled.

@@

*Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$*

*Initially my thought was to raw data only after aligning the resolution: Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem. After processing them and align them together they become the "raw data"*

Once variables are collected, they can have different spatial resolution and temporal granularity. Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data that contain the spatial, temporal, and multivariate aspect: the spatial component $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal component $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$. Expanding on the multivariate aspect, the data can be written as  $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$

@@

Temporal processing concerns the operations on the temporal aspect of the data. It can be represented by $f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))$, where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$.

- purpose is to aggregate across time, smooth?

An example of temporal processing is to aggregate the data, $x(s_i; t_j)$, into $k$-month sum. This can be used to reflect how the accumulated values change across time. In the drought index SPI, precipitation can be aggregated at different scales to reflect short-term, mid-term, and long-term drought severity. With $\psi = k$, the aggregated series can be written as $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)$ where $j^\prime$ is the time index after the aggregation.

@@

Sometimes, it can be beneficial to borrow information from nearby spatial locations or there is the need to combine raster data from satellite images with vector ground measures. Spatial aggregation can be written as $g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))$, where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$.

 - $\theta$ can be a constant or a derived from the data

 - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level.

**Variable transformation**

Variable transformation can be necessary for different reasons: 1) to change the shape of the variable, 2) to stablise the variance.

mention modelling? -> satisfy model assumption

A variable transformation changes the shape of the variable. It can be written as $h_{\tau}(x(\mathbf{s};\mathbf{t}))$, where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the potential parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$.
Variable transformation involve square root transformation, log transformation, and also include those non-linear transformation.

@@

Sometimes, scaling, as a specific type of variable transformation, is merged into dimension reduction as a pre-step, i.e. transform the covariance matrix into correlation matrix before applying the Principal Component Analysis (PCA), but it worth to be separated into its own step to make it explicit.
Its main different from variable transformation is that scaling does not change the shape of the data and it can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
@fig-scale-var-trans-compare shows a range of operations on a simulated variable. 
For example, In a z-score standardisation, $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$ are used to standardise the data by the its mean and standard deviation across all time and all space units
Another example is to scale the variable into a unit interval, which can be written as $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
While being a common technique in data pre-processing, scaling has an impact on the outliers of the data and analysts need to be cautious on checking its effect.

<!-- z-score standardising, min-max standardisation into [0, 1] or [0, 100], percentage change on the baseline -->
<!-- close to variable transformation step -->

@@

When multiple variables are collected in space and time to construct indices, they are often summarised with dimension reduction. Dimension reduction $x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})$,  where $p^* = 1, 2, \cdots, P^*$,  $p = 1, 2, \cdots, P$, and $P^* < P$. The most commonly used dimension reduction techniquePprincipal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF), can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the variance of the data given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.

@@

The purpose of normalising is for cross-comparison using the normal score.
This step can get criticism from analysts for being unnecessarily forcing the scale to the normal scale.
This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]$, where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$. 
Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.

>>>>

<!-- *constructing time series index should also be encapsulated in the framework* -->

An overview of the pipeline is given in @fig-pipeline-steps to illustrate the construction from raw data to the final indexes.
The available intermediate steps include temporal processing, spatial aggregation, variable transformation, scaling, dimension reduction, normalising, benchmarking, and simplification. 
These steps are not required to be arranged in a fixed order and not all the steps have to appear when constructing each index. 
The pipeline includes modules for operations in the spatial, temporal, and multivariate aspect of the data, as well as modules for comparing and communicating indexes (benchmarking and simplification).
While the pipeline starts with the raw data, steps prior to this, i.e. define a useful concept to construct the index, collect high quality data relevant for measuring the concept, are also cricial to the sucess of an index.
The framework proposed in this paper assumes the objective of an index has been defined and relevant variables have been collected and quality controlled.

@@

<!-- *Discussion: whether we should allow raw data to have different spatial/ temporal resolution: This would require the notation $\mathbf{s}$ and $\mathbf{t}$ to depend on $p$* -->

<!-- *Initially my thought was to raw data only after aligning the resolution: Another section on original data directly downloaded, can have different spatial resolution, temporal granularity, data quality problem. After processing them and align them together they become the "raw data"* -->

<!-- Once variables are collected, they can have different spatial resolution and temporal granularity.  -->

Let $\mathbf{x}(\mathbf{s};\mathbf{t})$ denote the raw data with spatial, temporal, and multivariate aspect: the spatial dimension $\mathbf{s} =  (s_1, s_2, \cdots, s_n)^\prime$ is defined in the 2D space: $\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2$, the temporal dimension $\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime$ is defined in the 1D space: $\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}$. When more than one variable are invovled, the multivariate data can also be written as: $\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime$.

@@

While many variables are collected separately at each time period, the underlying phenomenon, characterised by the index, is continuous. 
For example, precipitation can be measured daily or monthly, a meteorological drought is characterised by a prelonged lack of precipitation. 
Temporal proocessing is a step in the pipeline that includes operations in the temporal dimension and it can be generally represented by $$f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),$$ where $\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}$ is the parameters associated with the temporal operation and $d_{\psi}$ is the number of parameter of $\psi$. The purpose of temporal aggregation can be to aggregate variable across time or to borrow information from nearby time periods to smooth or adjust the current value. The drought index SPI is constructed from the aggregated precipitation with a time scale parameter $\psi = k$. The aggregation step in SPI can be written as $x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)$ where $j^\prime$ is the time dimension after the aggregation.

@@

Sometimes, information from neighbouring spatial locations can also be borrowed when constructing indexes. 
This involves interpolating data from observed locations to the unobserved and aggregating data into different levels.
In the pipeline, spatial aggregation can be written as $$g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),$$ where $\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}$ is the spatial operation parameters and $d_{\theta}$ is the number of parameter of $\theta$. [GIVEN AN EXAMPLE]

<!-- $x(s_i^\prime;t_j) = \sum_{i \in \Omega}x(s_i; t_j)$ -->

<!-- Another common use of spatial processing is to integrate satellite imagaries with ground measures, also merge satellite imagaries with different resolutions.  -->

 <!-- - $\theta$ can be a constant or a derived from the data -->

 <!-- - 3 by 3 aggregation, another purpose: the spatial grid is too fine, aggregate to higher level. -->

**Variable transformation **

Variable transformation and scaling, in the next section, are both variable pre-processing steps to prepare the data with desire statistical property to work with. This can be a more stablised variance, a normal distribution that satisfies modelling assumptions, or a data range that required by the algorithms.
Variable tranformation is a general notion of a functional transformation on the variable: $$h_{\tau}(x(\mathbf{s};\mathbf{t})),$$ where $\tau \in T \subseteq \mathbb{R}^{d_{\tau}}$ is the parameter in the transformation, if any, and $d_{\tau}$ is the number of parameter of $\tau$. Typical examples of variable transformation are log, quadratic, square root, and box-cox transformation.

While scaling can be seen as a specific type of variable tranformation, it is separated into own step to make the scaling operation explicit in the pipeline.
The key difference between the two steps is that variable transformation typically changes the shape of the data while scaling only changes the data scale and can usually be written in the form of $$[x(s_i;t_j) - \alpha]/\gamma.$$
For example, A z-score standardisation can be written in the above form with $\alpha = \bar{x}(s; t)$ and $\gamma = \sigma(s; t)$, a min-max standardisation uses $\alpha = \min[x(s_i, t_j)]$ and $\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]$.
@fig-scale-var-trans-compare shows a collection of variable pre-processing operations and uses color to differentiate whether the operation is a variable transformation or a scaling step. All the scaling operations in green results in the same shape as the original data.

@@

<!-- While being a common technique in data pre-processing, scaling has an impact on the outliers of the data and analysts need to be cautious on checking its effect. -->

@@

While multivariate information can be collected on a concept, they need to be summarised into a single number for the final index. Dimension reduction step $$x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),$$ where $p^* = 1, 2, \cdots, P^*$ and $p = 1, 2, \cdots, P$, reduces the variable dimension from $P$ to $P^*$. The most commonly used dimension reduction technique Principal Component Analysis (PCA), also called Empirical Orthogonal Function (EOF) in earth science, can be written as $x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})$ where $\lambda_p$ is the loading of the PC1, derived from maximising the data variance given the constraint $\sum_{p=1}^P\lambda_p^2 = 1$.

@@

The purpose of normalising is for cross-comparison using the normal score.
This step can get criticism from analysts for being unnecessarily forcing the scale to the normal scale.
This step is usually in the end of the pipeline when the data have been processed into a univariate series. 
When the data are not distributed normally, a distributed is usually fitted before converting to the normal score via the normal reverse CDF function: $\Phi^{-1}(.)$: $$\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))],$$ where $\eta \in H \subseteq \mathbb{R}^{d_{\eta}}$ is the distribution parameter and $d_{\eta}$ is the number of parameter of $\eta$. 
Sometimes, to account for the monthly variation in the data, the fitting is done separately for each month and the noramlising step can be written as $\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]$, where $j^*$ is all the indexes that satisfy $j^* \mod 12 = m$ for each $m = 0, 1,\cdots, 11$.

### 2022-11-23 paper/paper.tex
more drafting on the pipeline steps

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

\emph{constructing time series index should also be encapsulated in the
framework}

Figure~\ref{fig-pipeline-steps} presents the index construction pipeline
from the raw data to the final indexes. The available intermediate steps
include temporal processing, spatial aggregation, variable
transformation, scaling, dimension reduction, normalising, benchmarking,
and simplification. These steps do not have to be arranged in sequence
and not all the steps are required to present in every index
construction. Spatial aggregation and temporal processing are operations
of the data in space and time Variable transformation, scaling, and
dimension reduction concerns operations on the multivariate aspect of
the data. Normalisation, benchmarking, and simplification can be viewed
as the ``post-processing'' of indexes, which addresses the
cross-comparison across different indexes and time, and index
communication.

\begin{itemize}
\tightlist
\item
  divide into pre and post-processing
\end{itemize}

More steps are involved before the mathematical construction of indexes.
This includes 1) defining a concept of interest, being useful for
decision making and 2) finding the relevant and measurable variables to
characterise the concept. These choices define the scope of the indexes
and the data availability for constructing an index but not the
construction itself, hence not included in the pipeline framework. The
framework proposed in this paper assumes the objective of an index has
been defined and relevant variables have been collected and quality
controlled.

@@

\emph{Discussion: whether we should allow raw data to have different
spatial/ temporal resolution: This would require the notation
\(\mathbf{s}\) and \(\mathbf{t}\) to depend on \(p\)}

\emph{Initially my thought was to raw data only after aligning the
resolution: Another section on original data directly downloaded, can
have different spatial resolution, temporal granularity, data quality
problem. After processing them and align them together they become the
``raw data''}

Once variables are collected, they can have different spatial resolution
and temporal granularity. Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\)
denote the raw data that contain the spatial, temporal, and multivariate
aspect: the spatial component
\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
temporal component \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
defined in the 1D space:
\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). Expanding on the
multivariate aspect, the data can be written as
\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\)

@@

Temporal processing concerns the operations on the temporal aspect of
the data. It can be represented by
\(f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t}))\), where
\(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
associated with the temporal operation and \(d_{\psi}\) is the number of
parameter of \(\psi\).

\begin{itemize}
\tightlist
\item
  purpose is to aggregate across time, smooth?
\end{itemize}

An example of temporal processing is to aggregate the data,
\(x(s_i; t_j)\), into \(k\)-month sum. This can be used to reflect how
the accumulated values change across time. In the drought index SPI,
precipitation can be aggregated at different scales to reflect
short-term, mid-term, and long-term drought severity. With \(\psi = k\),
the aggregated series can be written as
\(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)\)
where \(j^\prime\) is the time index after the aggregation.

@@

Sometimes, it can be beneficial to borrow information from nearby
spatial locations or there is the need to combine raster data from
satellite images with vector ground measures. Spatial aggregation can be
written as \(g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t}))\), where
\(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the spatial
operation parameters and \(d_{\theta}\) is the number of parameter of
\(\theta\).

\begin{itemize}
\item
  \(\theta\) can be a constant or a derived from the data
\item
  3 by 3 aggregation, another purpose: the spatial grid is too fine,
  aggregate to higher level.
\end{itemize}

\textbf{Variable transformation}

Variable transformation can be necessary for different reasons: 1) to
change the shape of the variable, 2) to stablise the variance.

mention modelling? -\textgreater{} satisfy model assumption

A variable transformation changes the shape of the variable. It can be
written as \(h_{\tau}(x(\mathbf{s};\mathbf{t}))\), where
\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the potential
parameter in the transformation, if any, and \(d_{\tau}\) is the number
of parameter of \(\tau\). Variable transformation involve square root
transformation, log transformation, and also include those non-linear
transformation.

@@

Sometimes, scaling, as a specific type of variable transformation, is
merged into dimension reduction as a pre-step, i.e.~transform the
covariance matrix into correlation matrix before applying the Principal
Component Analysis (PCA), but it worth to be separated into its own step
to make it explicit. Its main different from variable transformation is
that it can be written in the form of \[[x(s_i;t_j) - \alpha]/\gamma.\]
For example, In a z-score standardisation, \(\alpha = \bar{x}(s; t)\)
and \(\gamma = \sigma(s; t)\) are used to standardise the data by the
its mean and standard deviation across all time and all space units
Another example is to scale the variable into a unit interval, which can
be written as \(\alpha = \min[x(s_i, t_j)]\) and
\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\). While being a common
technique in data pre-processing, scaling has an impact on the outliers
of the data and analysts need to be cautious on checking its effect.

@@

When multiple variables are collected in space and time to construct
indices, they are often summarised with dimension reduction. Dimension
reduction
\(x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t})\),
where \(p^* = 1, 2, \cdots, P^*\), \(p = 1, 2, \cdots, P\), and
\(P^* < P\). The most commonly used dimension reduction
techniquePprincipal Component Analysis (PCA), also called Empirical
Orthogonal Function (EOF), can be written as
\(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
where \(\lambda_p\) is the loading of the PC1, derived from maximising
the variance of the data given the constraint
\(\sum_{p=1}^P\lambda_p^2 = 1\).

@@

The purpose of normalising is for cross-comparison using the normal
score. This step can get criticism from analysts for being unnecessarily
forcing the scale to the normal scale. This step is usually in the end
of the pipeline when the data have been processed into a univariate
series. When the data are not distributed normally, a distributed is
usually fitted before converting to the normal score via the normal
reverse CDF function: \(\Phi^{-1}(.)\):
\(\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))]\), where
\(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
parameter and \(d_{\eta}\) is the number of parameter of \(\eta\).
Sometimes, to account for the monthly variation in the data, the fitting
is done separately for each month and the noramlising step can be
written as \(\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]\), where
\(j^*\) is all the indexes that satisfy \(j^* \mod 12 = m\) for each
\(m = 0, 1,\cdots, 11\).

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi

@@

An overview of the pipeline is given in Figure~\ref{fig-pipeline-steps}
to illustrate the construction from raw data to the final indexes. The
available intermediate steps include temporal processing, spatial
aggregation, variable transformation, scaling, dimension reduction,
normalising, benchmarking, and simplification. These steps are not
required to be arranged in a fixed order and not all the steps have to
appear when constructing each index. The pipeline includes modules for
operations in the spatial, temporal, and multivariate aspect of the
data, as well as modules for comparing and communicating indexes
(benchmarking and simplification). While the pipeline starts with the
raw data, steps prior to this, i.e.~define a useful concept to construct
the index, collect high quality data relevant for measuring the concept,
are also cricial to the sucess of an index. The framework proposed in
this paper assumes the objective of an index has been defined and
relevant variables have been collected and quality controlled.

@@

Let \(\mathbf{x}(\mathbf{s};\mathbf{t})\) denote the raw data with
spatial, temporal, and multivariate aspect: the spatial dimension
\(\mathbf{s} = (s_1, s_2, \cdots, s_n)^\prime\) is defined in the 2D
space: \(\mathbf{s} \in \mathcal{D}_s \subseteq \mathbb{R}^2\), the
temporal dimension \(\mathbf{t} = (t_1, t_2, \cdots, t_J)^\prime\) is
defined in the 1D space:
\(\mathbf{t} \in \mathcal{D}_t \subseteq \mathbb{R}\). When more than
one variable are invovled, the multivariate data can also be written as:
\(\mathbf{x}(\mathbf{s}; \mathbf{t}) = (x_1(\mathbf{s}; \mathbf{t}), x_2(\mathbf{s}; \mathbf{t}), \cdots, x_P(\mathbf{s}; \mathbf{t}))^\prime\).

@@

While many variables are collected separately at each time period, the
underlying phenomenon, characterised by the index, is continuous. For
example, precipitation can be measured daily or monthly, a
meteorological drought is characterised by a prelonged lack of
precipitation. Temporal proocessing is a step in the pipeline that
includes operations in the temporal dimension and it can be generally
represented by \[f_{\mathcal{\psi}}(x(\mathbf{s};\mathbf{t})),\] where
\(\psi \in \Psi \subseteq \mathbb{R}^{d_{\psi}}\) is the parameters
associated with the temporal operation and \(d_{\psi}\) is the number of
parameter of \(\psi\). The purpose of temporal aggregation can be to
aggregate variable across time or to borrow information from nearby time
periods to smooth or adjust the current value. The drought index SPI is
constructed from the aggregated precipitation with a time scale
parameter \(\psi = k\). The aggregation step in SPI can be written as
\(x(s_i;t_{j^\prime}) = \sum_{j = j^\prime-k+1}^{j^\prime}x(s_i; t_j)\)
where \(j^\prime\) is the time dimension after the aggregation.

@@

Sometimes, information from neighbouring spatial locations can also be
borrowed when constructing indexes. This involves interpolating data
from observed locations to the unobserved and aggregating data into
different levels. In the pipeline, spatial aggregation can be written as
\[g_{\mathcal{\theta}}(x(\mathbf{s};\mathbf{t})),\] where
\(\theta \in \Theta \subseteq \mathbb{R}^{d_{\theta}}\) is the spatial
operation parameters and \(d_{\theta}\) is the number of parameter of
\(\theta\). {[}GIVEN AN EXAMPLE{]}

\textbf{Variable transformation }

Variable transformation and scaling, in the next section, are both
variable pre-processing steps to prepare the data with desire
statistical property to work with. This can be a more stablised
variance, a normal distribution that satisfies modelling assumptions, or
a data range that required by the algorithms. Variable tranformation is
a general notion of a functional transformation on the variable:
\[h_{\tau}(x(\mathbf{s};\mathbf{t})),\] where
\(\tau \in T \subseteq \mathbb{R}^{d_{\tau}}\) is the parameter in the
transformation, if any, and \(d_{\tau}\) is the number of parameter of
\(\tau\). Typical examples of variable transformation are log,
quadratic, square root, and box-cox transformation.

@@

While scaling can be seen as a specific type of variable tranformation,
it is separated into own step to make the scaling operation explicit in
the pipeline. The key difference between the two steps is that variable
transformation typically changes the shape of the data while scaling
only changes the data scale and can usually be written in the form of
\[[x(s_i;t_j) - \alpha]/\gamma.\] For example, A z-score standardisation
can be written in the above form with \(\alpha = \bar{x}(s; t)\) and
\(\gamma = \sigma(s; t)\), a min-max standardisation uses
\(\alpha = \min[x(s_i, t_j)]\) and
\(\gamma = \max[x(s_i, t_j)] - \min[x(s_i, t_j)]\).
Figure~\ref{fig-scale-var-trans-compare} shows a collection of variable
pre-processing operations and uses color to differentiate whether the
operation is a variable transformation or a scaling step. All the
scaling operations in green results in the same shape as the original
data.

{\centering \includegraphics{paper_files/figure-pdf/fig-scale-var-trans-compare-1.pdf}

\caption{\label{fig-scale-var-trans-compare}Comparison of scaling and
variable transformation.}

@@

While multivariate information can be collected on a concept, they need
to be summarised into a single number for the final index. Dimension
reduction step
\[x_{p^*}(\mathbf{s}; \mathbf{t}) \rightarrow x_p(\mathbf{s}; \mathbf{t}),\]
where \(p^* = 1, 2, \cdots, P^*\) and \(p = 1, 2, \cdots, P\), reduces
the variable dimension from \(P\) to \(P^*\). The most commonly used
dimension reduction technique Principal Component Analysis (PCA), also
called Empirical Orthogonal Function (EOF) in earth science, can be
written as
\(x_{p^*}(\mathbf{s}; \mathbf{t}) = \sum_{p = 1}^{P}\lambda_{p}x_p(\mathbf{s};\mathbf{t})\)
where \(\lambda_p\) is the loading of the PC1, derived from maximising
the data variance given the constraint \(\sum_{p=1}^P\lambda_p^2 = 1\).

@@

The purpose of normalising is for cross-comparison using the normal
score. This step can get criticism from analysts for being unnecessarily
forcing the scale to the normal scale. This step is usually in the end
of the pipeline when the data have been processed into a univariate
series. When the data are not distributed normally, a distributed is
usually fitted before converting to the normal score via the normal
reverse CDF function: \(\Phi^{-1}(.)\):
\[\Phi^{-1}[F_{\eta}(x(\mathbf{s}; \mathbf{t}))],\] where
\(\eta \in H \subseteq \mathbb{R}^{d_{\eta}}\) is the distribution
parameter and \(d_{\eta}\) is the number of parameter of \(\eta\).
Sometimes, to account for the monthly variation in the data, the fitting
is done separately for each month and the noramlising step can be
written as \(\Phi^{-1}[F_{\eta}^m(x(\mathbf{s};t_{j^*}))]\), where
\(j^*\) is all the indexes that satisfy \(j^* \mod 12 = m\) for each
\(m = 0, 1,\cdots, 11\).
